---
title: "Explore_Peak_Discharge"
author: "KEC"
date: "2026-03-17"
output: html_document
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("dplyr")
library("tidyverse")
library("ggplot2")
library("sf")
library("mapview")
library("dataRetrieval")
library("nhdplusTools")
library("rvest")
library("ggthemes")
library("janitor")
library("data.table")
library("ggpubr")
library("cdssr")
library("broom")
library("here")
library("GGally")
library("soilDB")
library("purrr")
library("ggcorrplot")
library("mgcv")
library("climateR")
library("gratia")
library("plotly")
library("terra")
library("tidyr")
library("randomForest")


```

# Define get_snotel function

This is a wrapper around soilDB::fetchSCAN() that retrieves SNOTEL/SCAN station 
data either by spatial intersection with an AOI polygon or by explicit site 
number. Data are fetched in parallel using furrr/future and assembled into a 
single sf object to speed things up.

```{r get_snotel,  message=FALSE, warning=FALSE, results='hide'}
#' Pull in SNOTEL site data for an AOI <- imports SNOTEL and SCAN site data
#'
#' @param aoi An sf polygon/multipolygon object.
#' @param dist The distance (in the same units as the aoi object) to buffer around the AOI
#' @param years The years to pull data for. Default set to "all" which pulls all available years for each site, but can be a string of years
#' @param timeseries either 'Daily' or 'Hourly
#' @param save Whether to save the resulting file (TRUE/FALSE)
#' @param path If `save == TRUE`, the file path to save to
#' @return An sf object of weather stations within given spatial boundary and attributes

getSNOTEL <- function(aoi,
                      site_nums = NULL,
                      dist = 0,
                      years = "all",
                      timeseries = "Daily",
                      workers = 4,        # new: number of parallel workers
                      save = FALSE,
                      path = NULL) {

  # --- site metadata setup (unchanged) ---
  data('SCAN_SNOTEL_metadata', package = 'soilDB')
  sites <- st_as_sf(SCAN_SNOTEL_metadata,
                    coords = c("Longitude", "Latitude"),
                    crs = 4326)

  if (!is.null(aoi)) {
    aoi <- aoi %>% sf::st_buffer(dist = dist)
    site_aoi <- sites %>%
      st_transform(st_crs(aoi)) %>%
      st_filter(aoi) %>%
      distinct(Site, .keep_all = TRUE) %>%
      sf::st_join(aoi) %>%
      dplyr::select(Name, Site, State, County = countyName,
                    beginDate, endDate, Elevation_ft, HUC, geometry) %>%
      dplyr::mutate(
        start_year = ifelse(years %in% c("all", "All"), year(beginDate),  min(as.numeric(years))),
        end_year   = ifelse(years %in% c("all", "All"), year(Sys.Date()), max(as.numeric(years)))
      )
  } else {
    site_aoi <- sites %>%
      distinct(Site, .keep_all = TRUE) %>%
      dplyr::select(Name, Site, State, County = countyName,
                    beginDate, endDate, Elevation_ft, HUC, geometry) %>%
      dplyr::filter(Site %in% site_nums) %>%
      dplyr::mutate(
        start_year = ifelse(years %in% c("all", "All"), year(beginDate),  min(as.numeric(years))),
        end_year   = ifelse(years %in% c("all", "All"), year(Sys.Date()), max(as.numeric(years)))
      )
  }

  siteID <- unique(site_aoi$Site)

  # --- Parallel fetch (replaces the for loop) ---
  library(furrr)
  library(progressr)

  plan(multisession, workers = workers)

  with_progress({
    p <- progressor(along = siteID)

    dat <- future_map(siteID, function(id) {
      p(sprintf("Fetching site %s", id))

      site_meta <- site_aoi[site_aoi$Site == id, ]

      tryCatch(
        soilDB::fetchSCAN(
          site.code  = id,
          year       = site_meta$start_year:site_meta$end_year,
          timeseries = timeseries
        ) %>%
          map(~ mutate_all(., as.character)) %>%
          bind_rows() %>%
          as_tibble() %>%
          dplyr::select(-(Name:pedlabsampnum)),

        error = function(e) {
          message("Site ", id, " failed: ", e$message)
          NULL
        }
      )
    }, .options = furrr_options(seed = TRUE))
  })

  plan(sequential)

  # --- Assemble final output (unchanged) ---
  final_data <- bind_rows(dat) %>%
    left_join(site_aoi, by = "Site") %>%
    st_as_sf() %>%
    clean_names() %>%
    mutate(
      across(any_of(c("value", "elevation_ft", "depth", "water_year", "start_year", "end_year")), as.numeric),
      across(any_of(c("date", "datetime")), as.Date)
    )

  if (save == TRUE) {
    st_write(final_data, paste0(path, "/SNOTEL.shp"), append = FALSE)
  }

  return(final_data)
}

```



# uCLP Network

Read in uCLP Sonde Locations, set bounding box as area of interest, and
plot.

```{r uclp_network, message=FALSE, warning=FALSE}

sonde_locs <- read.csv("data/raw/spatial/metadata/sonde_location_metadata.csv") %>%
  dplyr::mutate(
    lat = str_split(lat_long, pattern = ",", simplify = TRUE)[,1] %>% as.numeric(),
    lng = str_split(lat_long, pattern = ",", simplify = TRUE)[,2] %>% as.numeric()) %>%
  st_as_sf(coords = c("lng","lat"), crs = 4326)

aoi <- sonde_locs %>%
  st_bbox() %>%
  st_as_sfc() %>%
  st_as_sf()

# Plot
mapviewOptions(fgb = FALSE, georaster = FALSE)
mapview(sonde_locs,
        zcol = "Site") +
  mapview(aoi, alpha.regions = 0)

```

# USGS

Download all USGS discharge data within area of interest that spans sonde locs.
This uses the new(ish) USGS Water Data API function calls. 

```{r usgs_flow_gauges}

# Define search area using aoi
bbox_vals <- sf::st_bbox(aoi) %>%
  as.numeric() %>% 
  round(digits = 5)

# Retrieve monitoring locations and calculate distance to water sources
usgs_sw <- read_waterdata_monitoring_location(bbox = bbox_vals, site_type = "Stream") %>%
  st_transform(crs = st_crs(aoi)) %>%
  st_filter(aoi) 

# View USGS SW stations
mapview(usgs_sw,
        layer.name = "USGS Flow Gauges")
```


Download daily discharge (parameter 00060, cfs) for all gauges found above. 
Records are filtered to post-1980, joined to station metadata, and a 7-day 
rolling mean (smooth_discharge) is calculated for downstream peak calculations.
Day-of-year and water year are also added. Results are plotted with both raw 
and smoothed time series for all stations.

```{r usgs_discharge, warnings = FALSE}

# Retrieve Daily Discharge Data for all stations in AOI
usgs_discharge <- read_waterdata_daily(
  monitoring_location_id = usgs_sw$monitoring_location_id,
  parameter_code = "00060" # Discharge, cubic feet per second
) %>%
  # filter to sites with data after 1980
  dplyr::filter(time > as.Date("1980-01-01"))

# Refine station list based on successful/filtered data retrieval
usgs_sw <- usgs_sw %>%
  dplyr::filter(monitoring_location_id %in% unique(usgs_discharge$monitoring_location_id))

# Join station metadata to discharge records
tidy_discharge <- usgs_discharge %>%
  left_join(usgs_sw %>% st_drop_geometry() %>%
              dplyr::select(monitoring_location_id, 
                            monitoring_location_name, 
                            altitude, 
                            agency_code),
            by = "monitoring_location_id") %>%
  dplyr::group_by(monitoring_location_id, monitoring_location_name) %>%
  dplyr::mutate(doy = yday(time),
                smooth_discharge = frollmean(value, n = 7),
                discharge = value,
                date = time,
                water_year = calcWaterYear(date),
                station_name = monitoring_location_name) %>%
  dplyr::ungroup() %>%
  dplyr::select(date, 
                water_year,
                monitoring_location_id, 
                station_name, 
                altitude, 
                agency_code, 
                discharge, 
                smooth_discharge, 
                unit_of_measure)

# Plot all stations
ggplot(tidy_discharge, 
       aes(x = date, color = station_name)) +
  #geom_point() +
  geom_line(aes(y = smooth_discharge)) +
  geom_line(aes(y = discharge), alpha = 0.2) +
  theme_bw() +
  labs(y = "Discharge (cfs)") +
  theme(legend.position = "none") +
  facet_wrap(~station_name, scales = "free")
```

# Select Gauge and peak timing

Subset to the Fort Collins gauge (Poudre at mouth of canyon), then for each 
water year identify the single day of peak smoothed discharge during the 
Jan–Jul snowmelt window. This gives one row per water year with the day-of-year
of the annual flow peak. The NHDPlus watershed polygon for this gauge is also 
retrieved for use in subsequent spatial analyses.

```{r select_discharge, warning = FALSE}
# Select ideal site
usgs_discharge_select <- tidy_discharge %>%
  dplyr::filter(grepl("FORT", station_name))

# Extract annual peak discharge
usgs_discharge_doy <- usgs_discharge_select %>%
  group_by(station_name) %>%
  mutate(
    global_start = min(date),
    global_end = max(date),
    doy = yday(date),
    year = year(date),
    month = month(date),
    water_year = calcWaterYear(date)
  ) %>%
  dplyr::filter(month %in% c(1,2,3,4,5,6,7)) %>%
  group_by(station_name, water_year) %>%
  # Filter for the row where value equals the maximum value of that group
  dplyr::filter(smooth_discharge == max(smooth_discharge, na.rm = TRUE)) %>%
  # Optional: If there are ties (two days with the same max), take the first one
  dplyr::slice(1) %>% 
  ungroup()

# Select just the station at the Poudre River Mouth moving forward
usgs_discharge_doy_s <- usgs_discharge_doy %>%
  dplyr::filter(grepl("FORT", station_name))


select_gage <- usgs_discharge_doy_s$monitoring_location_id[[1]] %>%
  nhdplusTools::map_nhdplus()

# Get watershed associated with gauge
watershed <- select_gage$basin


ggplot(usgs_discharge_doy_s, aes(x = water_year, y = doy)) +
  geom_point() +
  labs(x = "", y = "Peak Discharge DOY", title = usgs_discharge_doy_s$station_name[[1]]) +
  theme_bw()


#st_write(obj = watershed,
#         dsn = "data/raw/spatial/poudre_gage_watershed.shp")
  
```

# CDSSR

Read in CDSSR data from colorado dss. Here, we retrieve surface water stations 
and daily discharge time series from the Colorado Decision Support System via 
the cdssr package. Negative discharge values (instrument errors) are set to NA.
Stations that haven't reported in the last 3 days are filtered out to focus
on sites with real-time / current data.

Again, focus on the canyon mouth station ("MOUTH"). The CDSSR gauge loc is 
upstream  of major irrigation diversions compared to the USGS gauge. See the 
discharge comparison plot below for justification.

```{r CDSSR}

# Get stations in AOI
cdssr_stations <- get_sw_stations(aoi = aoi, radius = 15) %>%
  st_as_sf(coords = c("longitude", "latitude"), na.fail = FALSE, crs = 4326)

# Get discharge in AOI
cdssr_discharge <- cdssr_stations$station_num %>%
  purrr::map_dfr(~ get_sw_ts(station_number = .)) %>% 
         dplyr::left_join(cdssr_stations %>% 
            dplyr::select(station_num, station_name, 
                          county, water_district, more_information), 
            by = c("station_num")) %>%
  dplyr::mutate(value = ifelse(value < 0, NA, value))

cdssr_stations_new <- cdssr_stations %>% 
  filter(station_num %in% 
           (cdssr_discharge %>% filter(datetime > as.Date(Sys.Date() - days(3))) %>%
                  pull(station_num) %>%
                    unique()))

cdssr_tidy_discharge <- cdssr_discharge %>%
  select(c("date" = datetime,
           "monitoring_location_id" = station_num,
           station_name,
           "agency_code" = data_source,
           "discharge" = value)) %>%
  mutate(date = as.Date(date),
          water_year = calcWaterYear(date)) %>%
  group_by(water_year) %>%
  dplyr::mutate(smooth_discharge = frollmean(discharge, n = 7),
                unit_of_measure = "ft^3/s",
                altitude = NA)


cdssr_select_discharge <- cdssr_tidy_discharge %>%
  dplyr::filter(grepl("MOUTH", station_name))

ggplot(cdssr_select_discharge, #%>%
 #        dplyr::filter(datetime > as.Date(Sys.Date()) - years(1)), 
       aes(x = date, y = discharge, color = station_name)) +
  geom_line() +
  facet_wrap(~station_name, scales = "free") +
  theme_bw() +
  theme(legend.position = "none")

# Extract annual peak discharge
cdssr_discharge_doy <- cdssr_select_discharge %>%
  group_by(station_name) %>%
  mutate(
    global_start = min(date),
    global_end = max(date),
    doy = yday(date),
    year = year(date),
    month = month(date),
    water_year = calcWaterYear(date)
  ) %>%
  dplyr::filter(month %in% c(1,2,3,4,5,6,7)) %>%
  group_by(station_name, water_year) %>%
  # Filter for the row where value equals the maximum value of that group
  dplyr::filter(smooth_discharge == max(smooth_discharge, na.rm = TRUE)) %>%
  # Optional: If there are ties (two days with the same max), take the first one
  dplyr::slice(1) %>% 
  ungroup()

ggplot(cdssr_discharge_doy, aes(x = water_year, y = doy)) +
  labs(x = "", y = "Peak Discharge DOY", title = cdssr_discharge_doy$station_name[[1]]) +
  geom_point() +
  theme_bw()


```

# Compare cdssr and usgs

Plot of CDSSR and USGS discharge. Canyon mouth discharge is greater (and upstream)
implying less influence from diversions. This plot shows why using discharge at
the canyon mouth is essential.

```{r discharge_comp}

comb_dis <- usgs_discharge_select %>%
  select(date, station_name, discharge, water_year) %>%
  bind_rows(cdssr_select_discharge %>% select(date, station_name, discharge, water_year)) %>%
  filter(water_year > 1980)

ggplot(comb_dis, aes(x = date, y = discharge, color = station_name)) +
  geom_line() +
  theme_bw() +
  labs(color = "", x= "", y = "Discharge (cfs)")+
  theme(legend.position = "bottom")


```

# SNOTEL

Get SNOTEL data for all sites within AOI. Pull daily sensor records for three 
pre-selected SNOTEL sites in the Poudre headwaters (Joe Wright 1122, 
Long Draw 1123, Hourglass 551). Raw sensor data are widened, then melt (decline 
in SWE) is computed and accumulated at 3/7/14/28-day windows. Cumulative 
precipitation is differentiated to recover daily values.

We also calculate snotel water year summary statistics for predicting peak 
discharge and discharge timing. Note, we aim to make a prediction on April 1 
here, so all statistics for a given year need to be measurable by that time.

```{r get_SNOTEL}

# Search for sites given AOI
#snotel <- getSNOTEL(aoi = aoi,
#                    dist = 5000,
#                    years = "all",
#                    save = FALSE) %>%
#  clean_names() %>%
#  st_transform(4326)

# Since we already know the relevant sites
snotel_site_nums <- c("1122", "1123", "551")
snotel <- getSNOTEL(aoi = NULL,
                    site_nums = snotel_site_nums,
                    years = "all",
                    save = FALSE) %>%
  clean_names() %>%
  st_transform(4326)

# Get metadata
snotel_metadata <- SCAN_sensor_metadata(site.code = snotel_site_nums)

# Select unique sites for plotting locations
snotel_sites <- snotel %>%
  group_by(site) %>%
  dplyr::slice(1) %>% 
  ungroup()


# Create clean version of snotel data
snotel_tidy <- snotel %>%
  st_drop_geometry() %>%
  data.table() %>%
  dplyr::filter(sensor_id %in% 
                  c("PREC.I", "SNWD.I", "WTEQ.I", "TAVG.D", 
                    "TMIN.D", "TMAX.D", "SMS.I_2", "SMS.I_8", "SMS.I_20" )) %>%
  dplyr::select(date, name, value, sensor_id, water_year) %>%
  dplyr::group_by(date, name, sensor_id) %>%
  # Sometimes sensors have multiple values per day - summarize to daily with mean
  dplyr::summarize(value = mean(value, na.rm = TRUE),
                   water_year = mean(water_year, na.rm = TRUE), .groups = "keep") %>%
  dplyr::group_by(date, name) %>%
  pivot_wider(names_from = sensor_id, values_from = value) %>% 
  clean_names() %>% 
  dplyr::group_by(name) %>%
  dplyr::mutate(melt = c(NA, diff(wteq_i))) %>%
  dplyr::mutate(melt = ifelse(melt < 0, -melt, 0)) %>%
  dplyr::mutate(melt_3_day = frollsum(melt, 3, align = "left"), 
                melt_7_day = frollsum(melt, 7, align = "left"),
                melt_14_day = frollsum(melt, 14, align = "left"),
                melt_28_day = frollsum(melt, 28, align = "left"),
                water_year = calcWaterYear(date)) %>%
  group_by(name, water_year) %>%
  arrange(date) %>%
  mutate(
    cum_prec_i = prec_i,
    # Calculate daily precip from the cumulative 'prec_i' column
    # pmax(0, ...) ensures we don't get negative precip if the sensor is reset
    prec_i = pmax(0, prec_i - lag(prec_i, default = first(prec_i)))
  ) %>%
  ungroup()


# Calculate snotel water year summary statistics for predicting
# peak discharge and discharge timing. Note, we aim to make a prediction on
# April 1 here, so all statistics for a given year need to be measurable by that 
# time.

# 1. Calculate Pre-April Summary Metrics
annual_summary_base <- snotel_tidy %>%
  mutate(
    month = month(date),
    doy = yday(date),
    # Boolean flags for different seasonal windows
    is_monsoon = month %in% c(7, 8, 9),
    is_pre_apr = month > 9 | month < 4,
    is_apr1    = month == 4 & day(date) == 1
  ) %>%
  group_by(name, water_year) %>%
  summarize(
    
    # --- Peak & Melt Metrics (Oct-March window) ---
    peak_swe = if(any(is_pre_apr)) max(wteq_i[is_pre_apr], na.rm = TRUE) else NA,
    
    peak_swe_doy = if(any(is_pre_apr & !is.na(wteq_i))) {
        doy[is_pre_apr][which.max(wteq_i[is_pre_apr])]
      } else {
        NA_integer_
      },
      
    total_precip = if(any(is_pre_apr)) max(prec_i[is_pre_apr], na.rm = TRUE) else NA,
    max_daily_melt  = max(melt[is_pre_apr], na.rm = TRUE),
    max_7day_melt   = max(melt_7_day[is_pre_apr], na.rm = TRUE),
    precip_to_swe_ratio = if(any(!is.na(wteq_i[is_pre_apr])) && max(wteq_i[is_pre_apr], na.rm=T) > 0) {
      max(prec_i[is_pre_apr], na.rm = TRUE) / max(wteq_i[is_pre_apr], na.rm = TRUE)
    } else {
      NA_real_
    },
    swe_centroid_doy = if(any(!is.na(wteq_i[is_pre_apr])) && sum(wteq_i[is_pre_apr], na.rm=T) > 0) {
      sum(doy[is_pre_apr] * wteq_i[is_pre_apr], na.rm = TRUE) / sum(wteq_i[is_pre_apr], na.rm = TRUE)
    } else {
      NA_real_
    },
    
    # --- Pre-April Climate/Ripening ---
    early_melt_days     = sum(melt[is_pre_apr] > 0, na.rm = TRUE),
    mean_temp_pre_april = mean(tavg_d[is_pre_apr], na.rm = TRUE),
    
    
    # --- Soil Moisture Current Year (Oct-March) ---
    soil_moisture_2  = mean(sms_i_2[is_pre_apr], na.rm = TRUE),
    soil_moisture_8  = mean(sms_i_8[is_pre_apr], na.rm = TRUE),
    soil_moisture_20 = mean(sms_i_20[is_pre_apr], na.rm = TRUE),
    
    # --- Monsoon Metrics (July-Sept window) ---
    monsoon_total_precip = sum(prec_i[is_monsoon], na.rm = TRUE),
    monsoon_intensity    = max(prec_i[is_monsoon], na.rm = TRUE), 
    
    # Monsoon Dry Spell logic
    monsoon_dry_spell = {
      m_prec = prec_i[is_monsoon]
      m_prec = m_prec[!is.na(m_prec)] # Remove NAs first
      if(length(m_prec) == 0) {
        NA_real_ 
      } else {
        runs = rle(m_prec == 0)
        # Only take max if there ARE dry days (values == TRUE)
        if(any(runs$values)) max(runs$lengths[runs$values]) else 0
      }
    },
    
    # --- The April 1st Case ---
    # Now handled internally since we didn't filter the main DF
    swe_apr1 = wteq_i[is_apr1][1],
    
    .groups = "drop"
  )

# 2. Calculate Previous Year Soil Moisture and monsoon precip
prev_year_soil <- annual_summary_base %>%
  select(name, water_year, soil_moisture_2, soil_moisture_8, soil_moisture_20, 
         monsoon_total_precip, monsoon_intensity, monsoon_dry_spell) %>%
  mutate(water_year = water_year + 1) %>%
  rename_with(~paste0(., "_prev"), starts_with(c("soil_moisture", "monsoon")))

# 3. Final Join
annual_snotel_summary <- annual_summary_base %>%
  left_join(prev_year_soil, by = c("name", "water_year"))

print(annual_snotel_summary)


# plot map of sites
#  mapview(snotel_sites,
#          layer.name = "SNOTEL",
#          col.regions = "darkblue")


# Plot snotel data

ggarrange(
ggplot(snotel_tidy,
       aes(x = date, y =prec_i, color = name)) +
  geom_line() +
  theme_bw() +
  labs(y = "SWE"),
ggplot(snotel_tidy,
       aes(x = date, y = melt_28_day, color = name)) +
  geom_line() +
  theme_bw() +
  labs(y = "Melt", color = ""),
common.legend = TRUE, legend = "bottom")

```

```{r plot_site_locs}

mapview(watershed) +
   mapview(sonde_locs, col.regions = "seagreen") +
mapview(cdssr_stations_new, col.regions = "red") +
  mapview(usgs_sw, col.regions = "dodgerblue") +
  mapview(snotel_sites, col.regions = "darkblue")


```

# Climate data

Here, we bring in climate data (GridMET) for the entire watershed and summarize
to the daily average.


```{r gridmet}

climate_data <- climateR::getGridMET(AOI = watershed,
                                     varname = c("etr", "pr", "tmmx"),
                                     startDate = "2010-01-01",
                                     endDate = "2026-01-01",
                                     dryrun = FALSE)

clim_crs <- terra::crs(climate_data[[1]])

watershed_trans <- watershed %>% st_transform(crs = clim_crs)

climate_data <- climate_data %>%
            purrr::map(~terra::crop(., terra::vect(watershed_trans), mask = TRUE)) %>%          
            purrr::map_dfr(~ terra::as.data.frame(., xy = TRUE)) %>%
            data.table::data.table() %>%
            tidyr::pivot_longer(-c(x, y),
                                names_to = "var_temp",
                                values_to = "val") %>%
            tidyr::separate_wider_delim(var_temp, "_", names = c("var", "date")) %>%
            tidyr::drop_na(val) %>%
            dplyr::group_by(x, y, date) %>%
            tidyr::pivot_wider(names_from = "var", values_from = "val") %>%
            dplyr::mutate(date = as.Date(date)) %>%
            group_by(date) %>%
            dplyr::summarize(ppt_in = 0.0393701 * sum(pr, na.rm = TRUE),
                             tmmx = mean(tmmx - 273.15, na.rm = TRUE),
                             etr_in = mean(etr * 0.0393701, na.rm = TRUE))
                             

clim_stats <- climate_data %>%
  mutate(water_year = calcWaterYear(date),
    month = month(date),
    doy = yday(date),
    is_monsoon = month %in% c(7, 8, 9),
    is_pre_apr = month > 9 | month < 4) %>%
  group_by(water_year) %>%
  summarize(
    gm_precip_in_sum = if(any(is_pre_apr)) sum(ppt_in[is_pre_apr], na.rm = TRUE) else NA,
    gm_etr_in_mean = if(any(is_pre_apr)) mean(etr_in[is_pre_apr], na.rm = TRUE) else NA,
    gm_etr_in_max = if(any(is_pre_apr)) max(etr_in[is_pre_apr], na.rm = TRUE) else NA,
    gm_temp_c_max = if(any(is_pre_apr)) max(tmmx[is_pre_apr], na.rm = TRUE) else NA,
    .groups = "drop") %>%
  dplyr::filter(water_year > 2010)


```

Now, merge dischage and snotel data into a single dataframe "doy_data". Join the
CDSSR canyon mouth peak discharge timing/magnitude records with the SNOTEL 
annual summaries (Long Draw Reservoir) and GridMET climate stats into one df.

```{r doy_merge}


doy_data_2026 <- #usgs_discharge_doy_s %>% # Fort Collins Discharge
  cdssr_discharge_doy %>% # Canyon Mouth Discharge
  dplyr::rename(c("discharge_doy" = "doy",
                  "peak_discharge" = "smooth_discharge")) %>%
  left_join(annual_snotel_summary %>%
              dplyr::filter(name == 
                            #"Joe Wright"
                            "Long Draw Resv"
                            #"Hourglass Lake"
                              ),
            by = "water_year") %>%
  left_join(clim_stats, by = "water_year") %>%
  # Remove 1st year of data because of lagged variables
  filter(water_year >= 2011)  %>%
          select("water_year",
                 "discharge_doy",
                 "peak_discharge",
                 "peak_swe",
                 "peak_swe_doy",
                 "total_precip",
                 "max_daily_melt",
                 #"max_melt_doy",
                 "max_7day_melt",
                 #"max_7day_melt_doy",
                 "swe_apr1",
                 "monsoon_total_precip_prev",
                 "monsoon_intensity_prev",
                 "monsoon_dry_spell_prev",
                 #"soil_moisture_2",
                 #"soil_moisture_8",
                 #"soil_moisture_20",
                 "soil_moisture_2_prev",
                 "soil_moisture_8_prev",
                 #"soil_moisture_20_prev",
                 "mean_temp_pre_april",
                 "swe_centroid_doy",
                 "precip_to_swe_ratio",
                 "gm_precip_in_sum",
                 "gm_etr_in_mean",
                 "gm_etr_in_max",
                 "gm_temp_c_max"
                 ) 

doy_data <- doy_data_2026 %>%
  dplyr::filter(water_year < 2026)


```

Now, explore the data. We start by exploring linear relationships between 
dependent variables (peak_discharge and discharge_doy) and predictors. Compute 
Pearson correlations among all predictors and the two targets (discharge_doy 
and peak_discharge). Hierarchical clustering groups related variables together 
to reveal blocks of correlated predictors (e.g., SWE metrics, soil moisture).

```{r data_exploration,  fig.width=12, fig.height=10}

# Create a plot of all variables
#ggpairs(doy_data, columns = 2:ncol(doy_data), 
#        upper = list(continuous = wrap("cor", size = 3)),
#        mapping = aes(alpha = 0.5))


# Correlation Matrix
cor_matrix <- cor(doy_data %>% select(-water_year), 
                  use = "complete.obs")

ggcorrplot(cor_matrix, 
           hc.order = TRUE,         # Clusters similar variables together
           type = "lower",          # Only shows the bottom triangle (no duplicates)
           lab = TRUE,              # Shows the actual correlation coefficients
           lab_size = 3,
           method = "square", 
           colors = c("#E46726", "white", "#1f77b4"), # Prof. orange-blue scale
           title = "Predictor Correlation Matrix",
           ggtheme = theme_minimal())

```

Isolate and rank correlations between each predictor and the two targets 
(discharge timing and magnitude). Then fit a separate simple linear regression 
for every predictor and report R² — shows which individual variables carry the 
most linear explanatory power before multivariate models.

```{r target_cors, warning = FALSE}
# Calculate correlations specifically for targets
target_cors <- as.data.frame(cor_matrix) %>%
  rownames_to_column("Variable") %>%
  select(Variable, discharge_doy, peak_discharge) %>%
  filter(!Variable %in% c("discharge_doy", "peak_discharge")) %>%
  pivot_longer(cols = -Variable, names_to = "Target", values_to = "Correlation")

# Plot
ggplot(target_cors, aes(x = reorder(Variable, Correlation), y = Correlation, fill = Target)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("discharge_doy" = "#1f77b4", "peak_discharge" = "#E46726")) +
  labs(title = "Correlation of Predictors with Discharge Timing and Magnitude",
       x = "SNOTEL / Climate Variable",
       y = "Pearson Correlation (r)") +
  theme_minimal()


# Now, find individual R_squared using lm
predictors <- names(doy_data)[3:ncol(doy_data)]

# Map through predictors to find the best individual R-squared
doy_results <- predictors %>%
  map_df(~{
    formula <- as.formula(paste("discharge_doy ~", .x))
    mod <- lm(formula, data = doy_data)
    tibble(variable = .x, r_squared = summary(mod)$r.squared)
  }) %>%
  arrange(desc(r_squared))

print(doy_results)

mag_results <- predictors %>%
  map_df(~{
    formula <- as.formula(paste("peak_discharge ~", .x))
    mod <- lm(formula, data = doy_data)
    tibble(variable = .x, r_squared = summary(mod)$r.squared)
  }) %>%
  arrange(desc(r_squared))

print(mag_results)


```

Now, we use some other methods to explore predictor power, moving
outside assumptions of linear relationships.

First, we explore random forest for variable importance. This method
captures non-linear relationships and ranks variables by how much the
model's accuracy drops when they are removed.

-   %IncMSE (Mean Squared Error): Higher is better. If you "scramble"
    this variable, how much does the model's error jump? If a variable
    has a high %IncMSE, it means the model is heavily reliant on that
    variable for accurately predicting y.

-   IncNodePurity (Gini): Higher is more "pure." Measures how much the
    variable helps split the data into clean groups. This is a
    "structural" importance. High values here suggest the variable is a
    very consistent branching point in the forest's logic.
    

```{r timing_rf}
# Timing (discharge_doy)
rf_timing <- randomForest(discharge_doy ~ . - peak_discharge, 
                          data = doy_data %>% select(-water_year), importance = TRUE, na.action = na.omit)

importance_timing <- as.data.frame(importance(rf_timing)) %>%
  rownames_to_column(var = "Variable") %>%
  # Sort by %IncMSE (Percent Increase in Mean Squared Error)
  arrange(desc(`%IncMSE`))

print(importance_timing %>% head())

# IncMSE results indicate that discharge_doy is most influenced by:
# precip_to_swe_ratio, max_7day_melt_doy, soil_moisture_2_prev, max_7day_melt, max_daily_melt

ggplot(importance_timing, aes(x = reorder(Variable, `%IncMSE`), y = `%IncMSE`)) +
  geom_point(size = 3, color = "#003366") +
  geom_segment(aes(x = Variable, xend = Variable, y = 0, yend = `%IncMSE`), color = "grey") +
  coord_flip() +
  labs(title = "Variable Importance: Peak Discharge Timing",
       x = "Predictor",
       y = "% Increase in MSE (Importance)") +
  theme_minimal()


```
    

```{r magnitude_rf}

# Magnitude (peak_discharge)
rf_magnitude <- randomForest(peak_discharge ~ . - discharge_doy, 
                             data = doy_data %>% select(-water_year), importance = TRUE, 
                             na.action = na.omit)

# --- Magnitude Model (peak_discharge) ---
importance_magnitude <- as.data.frame(importance(rf_magnitude)) %>%
  rownames_to_column(var = "Variable") %>%
  arrange(desc(`%IncMSE`))

print(importance_magnitude %>% head())

# IncMSE results indicate that peak discharge is most influenced by:
# soil_moisture_2, total_precip, max_daily_melt, peak_swe, max_7day_melt_doy


ggplot(importance_magnitude, aes(x = reorder(Variable, `%IncMSE`), y = `%IncMSE`)) +
  geom_point(size = 3, color = "#003366") +
  geom_segment(aes(x = Variable, xend = Variable, y = 0, yend = `%IncMSE`), color = "grey") +
  coord_flip() +
  labs(title = "Variable Importance: Peak Discharge Magnitude",
       x = "Predictor",
       y = "% Increase in MSE (Importance)") +
  theme_minimal()


```


Since relationships are rarely linear (e.g., the effect of temperature
on snowmelt usually follows a curve rather than a straight line), we'll
test a Generalized Additive Model (GAM) that fits smooth splines to each
predictor. This is particularly useful for variables like
mean_temp_pre_april, where there might be a threshold effect.

```{r test_Gam_timing}

# 1. Model for Peak Timing (discharge_doy)
# We use 's()' for smooth terms. 
# 'k = 4' limits the "wiggliness" to prevent overfitting with small datasets.
#
#
# Previous analyses indicated the following were key variables:
# RF: soil_moisture_8_prev, monsoon_intensity_prev, max_daily_melt
# Corr: soil_moisture_8_prev, monsoon_dry_spell_prev, monsoon_intensity_prev


 # ADD LEAVE ONE OUT
gam_timing <- gam(discharge_doy ~ s(soil_moisture_8_prev, k = 4) +
                                  #s(monsoon_total_precip_prev, k = 4) +
                                  #s(gm_precip_in_sum, k = 4),
                                  s(peak_swe, k = 4),
                                  #s(monsoon_dry_spell_prev, k = 3),
                                  #s(gm_etr_in_mean, k = 3),
                                  ##s(monsoon_intensity_prev, k = 4),
                                  #s(soil_moisture_2_prev, k = 3),
                                  #s(soil_moisture_2, k = 4),
                  data = doy_data, 
                  method = "REML") # REML is more robust for small samples

# Print some gam stats
summary(gam_timing)

gam.check(gam_timing)

# Plot gam performance
appraise(gam_timing)

```

```{r test_gam_magnitude}
# 2. Model for Peak Magnitude (peak_discharge)
#
# Previous methods identified the following predictors:
# RF: monsoon_total_precip_prev, peak_swe, total_precip
# Corr: peak_swe, swe_apr1, total_precip, monsoon_total_precip_prev

gam_magnitude <- gam(peak_discharge ~ s(monsoon_total_precip_prev, k = 4) +
                                      s(peak_swe, k = 4),
                                      #s(gm_etr_in_mean, k = 4),
                                      #s(max_daily_melt, k = 4),
                     data = doy_data, 
                     method = "REML")

# Again print stats
summary(gam_magnitude)

# Plot performance
appraise(gam_magnitude)


```

Now, we'll develop n models using leave one out to create an ensemble
prediction.

```{r loo_ensemble_timing}
library(mgcv)
library(dplyr)

# 1. Remove rows with NAs
vars_needed <- c("discharge_doy", "soil_moisture_8_prev", "peak_swe")
doy_data_clean <- doy_data %>% filter(if_all(all_of(vars_needed), ~!is.na(.)))

n <- nrow(doy_data_clean)
n_dropped <- nrow(doy_data) - n
if (n_dropped > 0) message(n_dropped, " row(s) dropped due to missing values.")

# 2. Fit n LOO models, storing each one
#    Also capture each point's true out-of-sample prediction
loo_models <- vector("list", n)
loo_preds  <- numeric(n)

for (i in 1:n) {
  loo_models[[i]] <- gam(discharge_doy ~ s(soil_moisture_8_prev, k = 4) +
                                         s(peak_swe, k = 4),
                         data   = doy_data_clean[-i, ],
                         method = "REML")
  
  # True out-of-sample prediction: model i never saw point i
  loo_preds[i] <- predict(loo_models[[i]], newdata = doy_data_clean[i, ])
}

# 3. Build ensemble predictions by averaging across all n LOO models
#    Result is an n x n matrix: rows = data points, cols = LOO models
pred_matrix <- matrix(NA, nrow = n, ncol = n)

for (j in 1:n) {
  pred_matrix[, j] <- predict(loo_models[[j]], newdata = doy_data_clean)
}

ensemble_preds <- rowMeans(pred_matrix)

# 4. Compile results — two prediction columns for comparison
doy_results <- doy_data_clean %>%
  mutate(
    loo_pred       = loo_preds,       # each point predicted by the model that excluded it
    ensemble_pred  = ensemble_preds,  # average across all n LOO models
    loo_error      = discharge_doy - loo_pred,
    ensemble_error = discharge_doy - ensemble_pred
  )

# 5. Performance metrics for both approaches
ss_tot <- sum((doy_data_clean$discharge_doy - mean(doy_data_clean$discharge_doy))^2)

rmse_loo      <- sqrt(mean(doy_results$loo_error^2))
mae_loo       <-      mean(abs(doy_results$loo_error))
r2_loo        <- 1 - sum(doy_results$loo_error^2) / ss_tot

rmse_ensemble <- sqrt(mean(doy_results$ensemble_error^2))
mae_ensemble  <-      mean(abs(doy_results$ensemble_error))
r2_ensemble   <- 1 - sum(doy_results$ensemble_error^2) / ss_tot

# 6. Print results
cat("--- LOO Out-of-Sample Performance ---\n")
cat("RMSE: ", round(rmse_loo,      2), " days\n", sep = "")
cat("MAE:  ", round(mae_loo,       2), " days\n", sep = "")
cat("R²:   ", round(r2_loo,        3), "\n\n",    sep = "")

cat("--- Ensemble Performance (all n LOO models averaged) ---\n")
cat("RMSE: ", round(rmse_ensemble, 2), " days\n", sep = "")
cat("MAE:  ", round(mae_ensemble,  2), " days\n", sep = "")
cat("R²:   ", round(r2_ensemble,   3), "\n",      sep = "")
```

```{r loo_ensemble_magnitude}
library(mgcv)
library(dplyr)

# 1. Remove rows with NAs
vars_needed_mag <- c("peak_discharge", "monsoon_total_precip_prev", "peak_swe")
doy_data_clean_mag <- doy_data %>% filter(if_all(all_of(vars_needed_mag), ~!is.na(.)))

n_mag <- nrow(doy_data_clean_mag)
n_dropped <- nrow(doy_data) - n_mag
if (n_dropped > 0) message(n_dropped, " row(s) dropped due to missing values.")

# 2. Fit n LOO models, storing each one
loo_models_mag <- vector("list", n_mag)
loo_preds_mag  <- numeric(n_mag)

for (i in 1:n_mag) {
  loo_models_mag[[i]] <- gam(peak_discharge ~ s(monsoon_total_precip_prev, k = 4) +
                                              s(peak_swe, k = 4),
                             data   = doy_data_clean_mag[-i, ],
                             method = "REML")
  
  # True out-of-sample prediction: model i never saw point i
  loo_preds_mag[i] <- predict(loo_models_mag[[i]], newdata = doy_data_clean_mag[i, ])
}

# 3. Build ensemble predictions by averaging across all n LOO models
pred_matrix_mag <- matrix(NA, nrow = n_mag, ncol = n_mag)

for (j in 1:n_mag) {
  pred_matrix_mag[, j] <- predict(loo_models_mag[[j]], newdata = doy_data_clean_mag)
}

ensemble_preds_mag <- rowMeans(pred_matrix_mag)

# 4. Compile results
mag_results <- doy_data_clean_mag %>%
  mutate(
    loo_pred       = loo_preds_mag,
    ensemble_pred  = ensemble_preds_mag,
    loo_error      = peak_discharge - loo_pred,
    ensemble_error = peak_discharge - ensemble_pred
  )

# 5. Performance metrics
ss_tot_mag <- sum((doy_data_clean_mag$peak_discharge - mean(doy_data_clean_mag$peak_discharge))^2)

rmse_loo_mag      <- sqrt(mean(mag_results$loo_error^2))
mae_loo_mag       <-      mean(abs(mag_results$loo_error))
r2_loo_mag        <- 1 - sum(mag_results$loo_error^2) / ss_tot_mag

rmse_ensemble_mag <- sqrt(mean(mag_results$ensemble_error^2))
mae_ensemble_mag  <-      mean(abs(mag_results$ensemble_error))
r2_ensemble_mag   <- 1 - sum(mag_results$ensemble_error^2) / ss_tot_mag

# 6. Print results
cat("--- Magnitude LOO Out-of-Sample Performance ---\n")
cat("RMSE: ", round(rmse_loo_mag,      2), " cfs\n", sep = "")
cat("MAE:  ", round(mae_loo_mag,       2), " cfs\n", sep = "")
cat("R²:   ", round(r2_loo_mag,        3), "\n\n",   sep = "")

cat("--- Magnitude Ensemble Performance (all n LOO models averaged) ---\n")
cat("RMSE: ", round(rmse_ensemble_mag, 2), " cfs\n", sep = "")
cat("MAE:  ", round(mae_ensemble_mag,  2), " cfs\n", sep = "")
cat("R²:   ", round(r2_ensemble_mag,   3), "\n",     sep = "")
```

# Now, make predictions for 2026 using ensemble GAMs.

```{r timing_fit}

# Filter to just the 2026 row for prediction
new_data_2026 <- doy_data_2026 %>% filter(water_year == 2026)

# 1. Generate ensemble predictions for 2026
pred_2026_list <- lapply(loo_models, function(m) {
  predict(m, newdata = new_data_2026, se.fit = TRUE)
})

# Force to matrix even when new_data_2026 is a single row
fit_matrix_2026 <- matrix(sapply(pred_2026_list, function(p) as.numeric(p$fit)),
                          nrow = 1)
se_matrix_2026  <- matrix(sapply(pred_2026_list, function(p) as.numeric(p$se.fit)),
                          nrow = 1)

# Now apply works correctly
na_cols <- which(apply(fit_matrix_2026, 2, function(x) any(is.na(x))))
if (length(na_cols) > 0) {
  message(length(na_cols), " LOO model(s) excluded due to NA predictions.")
  fit_matrix_2026 <- fit_matrix_2026[, -na_cols, drop = FALSE]
  se_matrix_2026  <- se_matrix_2026[,  -na_cols, drop = FALSE]
}

# Ensemble point estimate and combined uncertainty
ensemble_fit_2026   <- mean(fit_matrix_2026)
mean_se2_2026       <- mean(se_matrix_2026^2)
intermodel_var_2026 <- var(as.numeric(fit_matrix_2026))
ensemble_se_2026    <- sqrt(mean_se2_2026 + intermodel_var_2026)

# 2. Build prediction row for 2026
pred_2026 <- new_data_2026 %>%
  mutate(
    predicted = ensemble_fit_2026,
    se        = ensemble_se_2026,
    lwr       = predicted - 1.96 * se,
    upr       = predicted + 1.96 * se
  )

# 3. Build historical fitted values from the ensemble (already computed)
hist_results <- doy_data_clean %>%
  mutate(
    predicted = ensemble_preds,   # from loo_ensemble_timing chunk
    error     = discharge_doy - predicted
  )

# 4. Plots
p1 <- ggplot() +
  # Historical observations
  geom_point(data = hist_results,
             aes(x = water_year, y = discharge_doy, color = "Observed"),
             alpha = 0.6) +
  # 2026 ensemble prediction
  geom_point(data = pred_2026,
             aes(x = water_year, y = predicted, color = "Predicted"),
             size = 3) +
  # 95% CI — combines model SE and inter-model spread
  geom_errorbar(data = pred_2026,
                aes(x = water_year, ymin = lwr, ymax = upr, color = "Predicted"),
                width = 0.2, linewidth = 0.8) +
  scale_color_manual(values = c("Observed" = "black", "Predicted" = "red")) +
  labs(
    title    = "Water Year 2026 Discharge Timing Prediction",
    subtitle = "Error bars represent 95% CI (model SE + inter-model variance)",
    x        = "Water Year",
    y        = "Peak Discharge (Day of Year)",
    color    = ""
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

p2 <- ggplot(hist_results, aes(x = discharge_doy, y = predicted)) +
  geom_point(alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "black",
              linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = -Inf, y = Inf,
           label = paste0("Ensemble R² = ", round(r2_ensemble, 2)),
           hjust = -0.3, vjust = 2, size = 5) +
  labs(
    title    = "LOO Ensemble GAM: Modeled vs. Observed Discharge DOY",
    subtitle = "Ensemble prediction averaged across all n LOO models",
    x        = "Observed (Day of Year)",
    y        = "Modeled (Day of Year)"
  ) +
  theme_bw()

ggarrange(p1, p2, align = "hv")
```

```{r magnitude_fit}
# Filter to just the 2026 row
new_data_2026_mag <- doy_data_2026 %>% filter(water_year == 2026)

# 1. Generate ensemble predictions for 2026 using magnitude LOO models
pred_2026_list_mag <- lapply(loo_models_mag, function(m) {
  predict(m, newdata = new_data_2026_mag, se.fit = TRUE)
})

# Force to matrix even when new_data_2026_mag is a single row
fit_matrix_2026_mag <- matrix(sapply(pred_2026_list_mag, function(p) as.numeric(p$fit)),
                              nrow = 1)
se_matrix_2026_mag  <- matrix(sapply(pred_2026_list_mag, function(p) as.numeric(p$se.fit)),
                              nrow = 1)

# Check for and remove NA columns
na_cols_mag <- which(apply(fit_matrix_2026_mag, 2, function(x) any(is.na(x))))
if (length(na_cols_mag) > 0) {
  message(length(na_cols_mag), " LOO model(s) excluded due to NA predictions.")
  fit_matrix_2026_mag <- fit_matrix_2026_mag[, -na_cols_mag, drop = FALSE]
  se_matrix_2026_mag  <- se_matrix_2026_mag[,  -na_cols_mag, drop = FALSE]
}

# Ensemble point estimate and combined uncertainty
ensemble_fit_2026_mag   <- mean(fit_matrix_2026_mag)
mean_se2_2026_mag       <- mean(se_matrix_2026_mag^2)
intermodel_var_2026_mag <- var(as.numeric(fit_matrix_2026_mag))
ensemble_se_2026_mag    <- sqrt(mean_se2_2026_mag + intermodel_var_2026_mag)

# 2. Build prediction row for 2026
pred_2026_mag <- new_data_2026_mag %>%
  mutate(
    predicted = ensemble_fit_2026_mag,
    se        = ensemble_se_2026_mag,
    lwr       = predicted - 1.96 * se,
    upr       = predicted + 1.96 * se
  )

# 3. Historical ensemble fitted values (from loo_ensemble_magnitude chunk)
hist_results_mag <- doy_data_clean_mag %>%
  mutate(
    predicted = ensemble_preds_mag,
    error     = peak_discharge - predicted
  )

# 4. Plots
p1_mag <- ggplot() +
  geom_point(data = hist_results_mag,
             aes(x = water_year, y = peak_discharge, color = "Observed"),
             alpha = 0.6) +
  geom_point(data = pred_2026_mag,
             aes(x = water_year, y = predicted, color = "Predicted"),
             size = 3) +
  geom_errorbar(data = pred_2026_mag,
                aes(x = water_year, ymin = lwr, ymax = upr, color = "Predicted"),
                width = 0.2, linewidth = 0.8) +
  scale_color_manual(values = c("Observed" = "black", "Predicted" = "red")) +
  labs(
    title    = "Water Year 2026 Peak Discharge Magnitude Prediction",
    subtitle = "Error bars represent 95% CI (model SE + inter-model variance)",
    x        = "Water Year",
    y        = "Peak Discharge (cfs)",
    color    = ""
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

p2_mag <- ggplot(hist_results_mag, aes(x = peak_discharge, y = predicted)) +
  geom_point(alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "black",
              linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = -Inf, y = Inf,
           label = paste0("Ensemble R² = ", round(r2_ensemble_mag, 2)),
           hjust = -0.3, vjust = 2, size = 5) +
  labs(
    title    = "LOO Ensemble GAM: Modeled vs. Observed Peak Discharge",
    subtitle = "Ensemble prediction averaged across all n LOO models",
    x        = "Observed (cfs)",
    y        = "Modeled (cfs)"
  ) +
  theme_bw()

ggarrange(p1_mag, p2_mag, align = "hv")
```
